Episodic memory: mental time travel or a quantum `memory wave' function?

### 2021-01-12 paper/old.tex
updated bibliography, draft of edits to address R1 comments, updated diff files

\section*{The shapes of remembered experiences}
The study by \cite{HeusEtal18c} described above suggests one way of characterizing the dynamic content of complex naturalistic stimuli like television episodes.  The sequence of topic vectors over the course of the episode traces out a \textit{topic trajectory} through a high-dimensional feature space.  The sequence of topic vectors for a given participant's recalls traces out an analogous trajectory.  This characterization provides a geometric framework for comparing the \textit{shapes} that reflect how naturalistic experiences unfold over time, to the shapes of how we remember those experiences later (Fig.~\ref{fig:trajectories}).  One can then use the geometric transformations needed to map an experience's trajectory onto the trajectory of its later remembering to characterize the quality and content of memory.

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/trajectory}
\caption{\textbf{The shape of an unfolding experience.}  \textbf{A. The trajectory through topic space taken by a television episode.} The topic trajectory taken by a 50 minute television episode, projected onto two dimensions.  Each dot indicates an event identified using a Hidden Markov Model; the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B. The trajectory through topic space taken by participants' recalls of the television episode.} The average two-dimensional trajectory captured by 17 participants' recalls of the episode, with the same format and coloring as the trajectory in Panel A. The arrows reflect the directions of the average transitions taken by participants whose trajectories crossed that part of topic space; blue arrows denote directions of reliable agreement across participants.  Figure adapted from \cite{HeusEtal18c}, where additional details may be found.}
\label{fig:trajectories}
\end{figure}

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal18c} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats} and~\ref{fig:reinstatement}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: to effectively communicate, we often need to convey the ``essence'' (gross overall shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

In their paper, \cite{HeusEtal18c} use the geometric shapes of experience trajectories (through word embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal18c} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  These are captured by low spatial frequency properties of the trajectory's shape.  Low-level details are characterized by (typically smaller-scale) within-event geometric patterns, which are captured by the high spatial frequency properties of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experiences over others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

@@

\section*{Acknowledgements}
I thank Luke Chang, Janice Chen, Paxton Fitzpatrick, Andrew Heusser, Christopher Honey, Caroline Lee, Talia Manning, Lucy Owen, Matthijs van der Meer, and Kirsten Ziman for feedback and scientific discussions. I also thank Janice Chen, Yuan Chang Leong, Christopher Honey, Chung Yong, Kenneth Norman, and Uri Hasson for sharing the data re-analyzed here~\citep{ChenEtal17}, and Paxton Fitzpatrick and Andrew Heusser for contributing analysis code that I adapted from a prior publication~\citep{HeusEtal18c}.  This work was supported in part by NSF EPSCoR Award Number 1632738. The content is solely my responsibility and does not necessarily represent the official views of my supporting organizations.

>>>>

\section*{The shapes of remembered experiences}
The study by \cite{HeusEtal21} described above suggests one way of characterizing the dynamic content of complex naturalistic stimuli like television episodes.  The sequence of topic vectors over the course of the episode traces out a \textit{topic trajectory} through a high-dimensional feature space.  The sequence of topic vectors for a given participant's recalls traces out an analogous trajectory.  This characterization provides a geometric framework for comparing the \textit{shapes} that reflect how naturalistic experiences unfold over time, to the shapes of how we remember those experiences later (Fig.~\ref{fig:trajectories}).  One can then use the geometric transformations needed to map an experience's trajectory onto the trajectory of its later remembering to characterize the quality and content of memory.

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/trajectory}
\caption{\textbf{The shape of an unfolding experience.}  \textbf{A. The trajectory through topic space taken by a television episode.} The topic trajectory taken by a 50 minute television episode, projected onto two dimensions.  Each dot indicates an event identified using a Hidden Markov Model; the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B. The trajectory through topic space taken by participants' recalls of the television episode.} The average two-dimensional trajectory captured by 17 participants' recalls of the episode, with the same format and coloring as the trajectory in Panel A. The arrows reflect the directions of the average transitions taken by participants whose trajectories crossed that part of topic space; blue arrows denote directions of reliable agreement across participants.  Figure adapted from \cite{HeusEtal21}, where additional details may be found.}
\label{fig:trajectories}
\end{figure}

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal21} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats} and~\ref{fig:reinstatement}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: to effectively communicate, we often need to convey the ``essence'' (gross overall shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

In their paper, \cite{HeusEtal21} use the geometric shapes of experience trajectories (through word embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal21} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  These are captured by low spatial frequency properties of the trajectory's shape.  Low-level details are characterized by (typically smaller-scale) within-event geometric patterns, which are captured by the high spatial frequency properties of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experiences over others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

@@

\section*{Acknowledgements}
I thank Luke Chang, Janice Chen, Paxton Fitzpatrick, Andrew Heusser, Christopher Honey, Caroline Lee, Talia Manning, Lucy Owen, Matthijs van der Meer, and Kirsten Ziman for feedback and scientific discussions. I also thank Janice Chen, Yuan Chang Leong, Christopher Honey, Chung Yong, Kenneth Norman, and Uri Hasson for sharing the data re-analyzed here~\citep{ChenEtal17}, and Paxton Fitzpatrick and Andrew Heusser for contributing analysis code that I adapted from a prior publication~\citep{HeusEtal21}.  This work was supported in part by NSF EPSCoR Award Number 1632738. The content is solely my responsibility and does not necessarily represent the official views of my supporting organizations.
